Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-24 astronum_2017/astronum-amrex.tex
wordsmithing

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro, designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent explosion.
Together this suite of codes allows us to address many problems in
stellar and nuclear astrophysics.  We describe some of the design
details, the current architecture of the code, and some applications
below.

@@

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is whether a deflagration
can transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
not clear whether it is possible to create a detonation in the thin
surface He layer.  For double degenerates, it is still unresolved
whether the burning takes place promptly or after some delay.  However
it proceeds, we need to avoid an accretion-induced collapse to a
neutron star.  For many of these investigations, we need to address
numerical issues such as whether it is possible to accurately model
the ignition of a detonation with the spatial resolution we can
attain.  These are some of the questions we seek to answer.

\subsection{XRBs}

X-ray bursts, the burning of accreted H/He on the surface of a neutron
star, can be important probes of neutron star structure.  Interpreting
observations requires that we understand what we are seeing, which can
be influenced by the products of the burning and how the burning
spreads across the star.  Many efforts have focused on different
aspects of these events.  One-dimensional models capture the
energetics well and inform us about the nucleosynthesis
\cite{woosley-xrb}.  Global models show the importance of rotation in
confining the burning \cite{SPIT_ETAL02}, while models inspired from
atmospheric science can explore the vertical structure
\cite{cavecchi:2012}.  However, the resolution differences from the
scale of the burning to a reasonable fraction of the neutron star
surface has prevented detailed explorations of the burning in resolved
calculations and how it feeds back on the flame structure and
propagation.  Algorithms and computing architectures are starting to
reach the point where we can span the gaps in spatial scales between
these calculations to provide a better understanding of the ignition
and propagation of the burning front, and the nucleosynthesis produced.

@@

These problems share common algorithmic requirements: strong coupling
between hydrodynamics and burning, support for a general equation of state,
self-gravity, including isolated boundary conditions, and long timescale
evolution for the convective phases.  All of these problems are inherently
three-dimensional, as turbulence, fluid instabilities, and rotation affect
the dynamics.  Conservation is important as well, suggesting approaches
which implement gravitational and rotation sources conservatively, and
methods for improving angular momentum conservation.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections require solving a variable
coefficient elliptic equations, which is done numerically using a
multigrid algorithm.  A number of recent advances in low Mach
number modeling~\cite{kleinpauluis,vasil:2013} have been
incorporated into \maestro.

\maestro\ has been applied to convection in the Chandrasekhar-mass
model for Type Ia supernovae (SNe Ia)
\cite{ZABNW:IV,wdconvect,wdturb}, the sub-Chandra model for SNe Ia
\cite{subchandra,subchandra2} X-ray bursts~\cite{xrb,xrb2,xrb3}, and
convection in massive stars \cite{ms_cc}.

@@

\castro\ has been applied to core-collapse
supernovae~\cite{castro-ccsne}, population III pair-instability
supernovae~\cite{castro-pairinstability}, and white dwarf mergers as a
model for SNe Ia~\cite{wdmergerI}.  Recent work has focused on
improving energy conservation with gravitational and rotational
sources and new integration methods that more accurately couple hydro
and reactions.  For \maestro\ simulations that evolve from the
subsonic regime to the sonic regime, we have demonstrated the ability
to restart the calculations in \castro\ to continue the evolution into
the sonic regime~\cite{scidac-petascale,malone:2014} (in this case,
for Chandra model SNe Ia).

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics github
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that the rates and integration strategy are decoupled, allowing us to
change the integration strategy for a given set of rates.  They are
also written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to create a set of community
microphysics solvers that can be used in a variety of nuclear
astrophysics codes, not just those discussed here.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
github\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
Additionally, we have mailing lists for discussions and asking for
help.  Several branches are used in our workflow.  New changes are put
into the {\tt development} branch in each repo.  Nightly regression
testing is used to ensure that no new bugs were introduced.  Once a
month, {\tt development} is merged into {\tt master}.  Finally, {\em all source
  files, model files, input parameters, etc.\ for any published
  science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in
paper acknowledgments.

@@

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF Titan.  The AMD processors in
Titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

@@

A parallel effort is porting our microphysics, in particular reaction
networks to GPUs.  Our strategy is to do the entire ODE integration on
the GPU, i.e., the data for a patch of zones is passed to the GPU, all
righthand size and Jacobian evaluations and the timestepping itself is
done on the GPU and once the burning in all zones is completed, we access the data as
needed on the CPU.  To enable this, we ported our workhorse ODE
integrator (VODE~\cite{vode}) to CUDA Fortran. This required extensive
rewrites of the internals of VODE, which was originally written using
Fortran 77 syntax unsupported by CUDA Fortran. Accelerating VODE with
CUDA Fortran has proven successful, and we now see significant
performance gains with the CUDA version of our reaction networks, even
for moderate sized networks (a 13-isotope standard network).
Figure~\ref{fig:cudaode} shows the speed-ups on a GPU vs.\ single CPU
core.

@@

We have described our suite of astrophysics codes built on the
\amrex\ block-structured adaptive mesh refinement framework.  These
codes were developed to model problems in stellar astrophysics
spanning from low speed convection to explosive burning.  A major
theme of the codes is the open development model, with the code
development done on github and all problem files needed to recreate
any science results freely available.  Future development efforts for
\maestro\ include higher-order hydrodynamics and time-integration and
rotation.  For \castro, we are investigating stronger coupling between
hydrodynamics and reactions, new solvers (including MHD), and
finishing the GPU port.

>>>>

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro, designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent explosion,
respectively.  Together this suite of codes allows us to address many
problems in stellar and nuclear astrophysics.  We describe some of the
design details, the current architecture of the code, and some
applications below.

@@

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots set up
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is whether a deflagration
can transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
not clear whether it is possible to create a detonation in the thin
surface He layer.  For double degenerates, it is still unresolved
whether the burning takes place promptly or after some delay.  However
it proceeds, we need to avoid an accretion-induced collapse to a
neutron star.  For many of these investigations, we need to address
numerical issues such as whether it is possible to accurately model
the ignition of a detonation with the spatial resolution we can
attain.  These are some of the questions we seek to answer.

\subsection{X-ray Bursts}

X-ray bursts --- the burning of accreted H/He on the surface of a
neutron star --- can be important probes of neutron star structure.
Interpreting observations requires that we understand what we are
seeing, which can be influenced by the products of the burning and how
the burning spreads across the star.  Many efforts have focused on
different aspects of these events.  One-dimensional models capture the
energetics well and inform us about the nucleosynthesis
\cite{woosley-xrb}.  Global models show the importance of rotation in
confining the burning \cite{SPIT_ETAL02}, while models inspired from
atmospheric science can explore the vertical structure
\cite{cavecchi:2012}.  However, the resolution differences from the
scale of the burning to a reasonable fraction of the neutron star
surface has prevented detailed explorations of the burning and how it
feeds back on the flame structure and propagation in resolved
calculations.  Algorithms and computing architectures are starting to
reach the point where we can span the gaps in spatial scales between
these calculations to provide a better understanding of the ignition
and propagation of the burning front, and the nucleosynthesis
produced.

@@

These problems share common algorithmic requirements: strong coupling
between hydrodynamics and burning, support for a general equation of state,
self-gravity, including isolated boundary conditions, and long timescale
evolution for the convective phases.  All of these problems are inherently
three-dimensional, as turbulence, fluid instabilities, and rotation affect
the dynamics.  Conservation is important as well, suggesting approaches
that implement gravitational and rotation sources conservatively, and
methods for improving angular momentum conservation.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections require solving a variable
coefficient elliptic equation, which is done numerically using a
multigrid algorithm.  A number of recent advances in low Mach number
modeling~\cite{kleinpauluis,vasil:2013} have been incorporated into
\maestro.

\maestro\ has been applied to convection in the Chandrasekhar-mass
model for SNe Ia \cite{ZABNW:IV,wdconvect,wdturb}, the sub-Chandra
model for SNe Ia \cite{subchandra,subchandra2},
XRBs~\cite{xrb,xrb2,xrb3}, and convection in massive stars
\cite{ms_cc}.

@@

\castro\ has been applied to core-collapse
supernovae~\cite{castro-ccsne}, population III pair-instability
supernovae~\cite{castro-pairinstability}, and white dwarf mergers as a
model for SNe Ia~\cite{wdmergerI}.  Recent work has focused on
improving energy conservation with gravitational and rotational
sources, and new integration methods that more accurately couple hydro
and reactions.  For \maestro\ simulations that evolve from the
subsonic regime to the sonic regime, we have demonstrated the ability
to restart the calculations in \castro\ to continue the evolution into
the sonic regime~\cite{scidac-petascale,malone:2014} (in this case,
for Chandra model SNe Ia).

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics GitHub
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that the rates and integration strategy are decoupled, allowing us to
change the integration strategy for a given set of rates.  They are
also written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to create a set of community
microphysics solvers that can be used in a variety of nuclear
astrophysics codes, not just those discussed here.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
GitHub\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
Additionally, we have mailing lists for discussions and asking for
help.  Several branches are used in our workflow.  New changes are put
into the {\tt development} branch in each repo.  Nightly regression
testing is used to ensure that no new bugs were introduced.  Once a
month, {\tt development} is merged into {\tt master}.  Finally, {\em all source
  files, model files, input parameters, etc.\ for any published
  science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in
paper acknowledgments.

@@

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a
pure hydro problem with a real EOS and a hydro + self-gravity problem
with a real EOS.  All tests were run on OLCF Titan.  The AMD
processors in Titan can be used in an 8 or 16 cores / node
configuration.  For all these runs, we ran in the 16 cores / node
mode.  All runs also used the PGI 17.7 compilers.  We see excellent
scaling behavior for the pure hydro problem.  For the self-gravity
problem, we scale well with a uniform grid, but with a factor of 4
refinement for the AMR level on top of this, we observe degraded
performance at higher core counts.  This is a challenging problem, as
only 3\% of the domain volume is refined.  For a given core count,
there are a number of different combinations of MPI ranks and OpenMP
threads we can use.  In general, with multilevel problems, we found
the best performance with fewer MPI ranks and more OpenMP threads.

@@

A parallel effort is porting our microphysics --- in particular
reaction networks --- to GPUs.  Our strategy is to do the entire ODE
integration on the GPU, i.e., the data for a patch of zones is passed
to the GPU, all righthand size and Jacobian evaluations and the
timestepping itself is done on the GPU, and once the burning in all
zones is completed, we access the data as needed on the CPU.  To
enable this, we ported our workhorse ODE integrator (VODE~\cite{vode})
to CUDA Fortran. This required extensive rewrites of the internals of
VODE, which was originally written using Fortran 77 syntax unsupported
by CUDA Fortran. Accelerating VODE with CUDA Fortran has proven
successful, and we now see significant performance gains with the CUDA
version of our reaction networks, even for moderate-sized networks (a
13-isotope standard network).  Figure~\ref{fig:cudaode} shows the
speed-ups on a GPU vs.\ single CPU core.

@@

We have described our suite of astrophysics codes built on the
\amrex\ block-structured adaptive mesh refinement framework.  These
codes were developed to model problems in stellar astrophysics
spanning from low speed convection to explosive burning.  A major
theme of the codes is the open development model, with the code
development done on GitHub and all problem files needed to recreate
any science results freely available.  Future development efforts for
\maestro\ include higher-order hydrodynamics and time-integration and
rotation.  For \castro, we are investigating stronger coupling between
hydrodynamics and reactions, new solvers (including MHD), and
finishing the GPU port.
